The Solar System: an Exercise in Numerical Integration and \\ \hrulefill\small{ FYS3150: Computational Physics }\hrulefill

### 2016-10-24 Project3/article/article.tex
nearly there..

\usepackage[]{graphicx}
\usepackage[]{hyperref}
\usepackage[]{physics}
\usepackage[]{listings}
\usepackage[utf8]{inputenc}
\usepackage[toc,page]{appendix}
\usepackage[dvipsnames]{xcolor}

@@

\title[Simulation of the Solar System]{The Solar System: an Exercise in Numerical Integration \\
  \hrulefill\small{ FYS3150: Computational Physics }\hrulefill}

@@

@@

The velocity Verlet method can also be represented by a (one-dimensional) recursive relation
\begin{align}
x_{n+1} &= x_n + v_ndt + \frac{1}{2}a_ndt^2 \\
v_{n+1} &= v_n + \frac{a_n + a_{n+1}}{2}dt.
\end{align}
These equations tell us that the algorithm is slightly more expensive compared to the Euler-Cromer scheme at $11n$. Here follows an implementation of the algorithm in C++.

@@

@@

\begin{figure}
	\centering
	\textbf{Multi-body model of the Solar System}
	\includegraphics[width=0.8\textwidth]{../figures/multibody.png}
	\caption{All planets of our solar system (including Pluto) \label{fig:multibody}}
\end{figure}

Finally for the pinnacle of this study, a model of the entire solar system!\footnote{Excluding moons and other satelites, but including Pluto.} Figure \ref{fig:multibody} shows a multi-body simulation representing the solar system, integrated using the velocity Verlet algorithm. One can see that the algorithm handles the problem quite well.

@@

Figure \ref{fig:perangle} shows the angel of the perihelion of Mercury calculated by equation \ref{eq:perangle} in the theory section. This plot is over a period of 30 Earth years and with a time step of $dt = 10^{-6}yrs$. One can see a clear trend in the angle of the perihelion as predicted.

@@

@@

Velocity Verlet for the win!! :D

@@

@@

>>>>

\usepackage[]{graphicx}
\usepackage[]{hyperref}
\usepackage[]{physics}
\usepackage[]{listings}
\usepackage[utf8]{inputenc}
\usepackage[toc,page]{appendix}
\usepackage[dvipsnames]{xcolor}
\usepackage{dirtytalk}

@@

\title[Simulation of the Solar System]{The Solar System: an Exercise in Numerical Integration and \\
  \hrulefill\small{ FYS3150: Computational Physics }\hrulefill}

@@

This study has three main goals. To show some of the advantages of object-oriented programming, to implement and compare different integration methods and to build a working model of the solar system.

@@

The velocity Verlet method can also be represented by a (one-dimensional) recursive relation
\begin{align}
x_{n+1} &= x_n + v_ndt + \frac{1}{2}a_ndt^2 \\
v_{n+1} &= v_n + \frac{a_n + a_{n+1}}{2}dt.
\end{align}
These equations tell us that the algorithm is slightly more expensive compared to the Euler-Cromer scheme at $11n$ FLOPS. Here follows an implementation of the algorithm in C++.

@@

In an arbitrary system, the entire system will usually gain momentum and drift off in some direction over time. One must usually correct for such drift be substracting total linear momenum. Because the velocity data from NASA is very good, this is unnecessary for our system. Total linear momemtum for the solar system is already zero from our frame of reference.

@@

\begin{figure}
	\centering
	\textbf{Multi-body model of the Solar System}
	\includegraphics[width=0.9\textwidth]{../figures/multibody.png}
	\caption{All planets of our solar system (including Pluto) \label{fig:multibody}}
\end{figure}

Finally for the pinnacle of this study, a model of the entire solar system!\footnote{Excluding moons and other satelites, but including Pluto.} Figure \ref{fig:multibody} shows a multi-body simulation representing the solar system, integrated using the velocity Verlet algorithm. One can see that the algorithm handles the problem quite well. We were even able to make an anmation of this system in MATLAB, which was a delightful exercise.

@@

Figure \ref{fig:perangle} shows the angel of the perihelion of Mercury calculated by equation \ref{eq:perangle} in the theory section. This plot is over a period of 30 Earth years and with a time step of $dt = 10^{-6}yrs$. One can see a clear trend in the angle of the perihelion as predicted. We get a slope of the perhelion angle versus number of orbist of approximately $10^{-6}$, for a century on earth Mercury makes $415.2091$ orbits, which gives us a perhelion precession of $\approx1.5''/100yrs$. This much lower than the expected precession of $43''/100yrs$.

@@

\subsection{Object Orientation}
This project has been a nice exercise in object oriented thinking. We have built a larger and larger system by adding more class instances and building subclasses such as different integrators. Object oriented programming is close to how we as humans usually thinks about all things in the world. A planet is something that has mass, a position, velocity and some other factors. However, when one learns how to program learns object orientation at a late stage of the learning process, when one has been focusing on a more procedural programming technique.

Most widely used programming languages today are object-oriented to a greater or lesser extent: Java, C++, Python, PHP, Ruby, Smalltalk and Common Lisp, to name a few. Critique of object orientation usually emphasizes that design and modelling comes at the expense of other important aspects like the actual computation and algorithms. The creator of Erlang, Joe Armstrong, is quoted as saying, \say{The problem with object-oriented languages is they've got all this implicit environment that they carry around with them. You wanted a banana but what you got was a gorilla holding the banana and the entire jungle}.

As fairly level-headed Norwegian patriots\footnote{A friendly reminder to the reader: the creators of the first object oriented languas, Simula, were Norwegian: Ole-Johan Dahl, Kristen Nygaard.} we cannot end the discussion about object orientation there, but rather by pointing out the advantages. Object Oriented Programming provides a clear modular structure for programs and makes it easy to maintain and modify existing code thereby reducing lowering programming ``cost''.

@@

We have seen that the velocity Verlet algorithm is much more precise than the Euler-Cromer method (figure \ref{tab:eulervsverlet}). In addition it appears that it may also be quicker (table \ref{tab:eulervsverlet}), which is a surprise since the Verlet algorithm seems to include more FLOPS than the Euler-Cromer method. This leads us to believe that the velocity Verlet may not be that much faster, at least we can conclude that the results are not statistically significant since we only have one observation per method per value for $N$. An improvement on this study would have been to conduct such a test. We propose larger sampling of code run time and, for simplicity, a paired Student's t-test as a start to answer this question\cite{student}.

The discussion of an Euler method versus a Verlet method is a rather big one. Both methods are quite quick cheap which is what you want, and they are usually the two methods of choice if one does not need entirely exact results, but results that are ``good enough''. A field where this discussion prominent is in physics engine design for video games. With more computing power one would probably go for a more sophisticated method like a foruth degree Runge-Kutta method\cite{physicsengine}.

It is worth metioning something about the symplecticity of the velocity Verlet method. As stated previously, a symplectic method will be good at simulating systems with energy conservation. We believe that the solar system has conserved energy, but our model of it may not be. Looking further into this matter we have found that both energy and angular momentum is conserved for the two-body Earth-Sun system. This would warrant further use of the velocity Verlet scheme. See appendix \ref{app:energy} for a sample readout of energies and angular momenta.

\subsection{Mercury}

We tried to compute the perihelion precessrion of Mercury. We found it to be $\approx1.5''/100yrs$ when it should have been much more, id est $43''/100yrs$. This error can stem from several sources. Most obvious is that we have done something wrong and have been unable to find where. Antoher likely source of uncertainty is that the method is that the integration method is not precise enough. We had problems decreasing the step size further, and extending the simulation to a full century because the job was too much for our hardware. One last probelm that must be mentioned is that Mercury's precession cannot be explained by the simple relativistic correction we are employing.

@@

\begin{thebibliography}{9}

\bibitem{erlang}
	Armstrong, Joe (2009),
	\emph{Coders at Work: Reflection on the Craft of Programming},
	Peter Seibel, ed.

\bibitem{student}
	Hazewinkel, Michiel, ed. (2001), 
	Student test,
	\emph{Encyclopedia of Mathematics},
	Springer.

\bibitem{physicsengine}
	Boesch, Florian (2010, August 28),
	\emph{Integration by Example - Euler vs Verlet vs Runge-Kutta},
	Retrieved October 24 2016 from \url{http://codeflow.org/entries/2010/aug/28/integration-by-example-euler-vs-verlet-vs-runge-kutta/},

\end{thebibliography}

@@

\section{Energy Read-out}
\label{app:energy}
\begin{lstlisting}[keywordstyle=\ttfamily, identifierstyle=\ttfamily]
Energies printed in C++:

Energies for Earth-Sun
EULER-CROMER
Angular momentum: 39.4784*m
Step: 1000000  E = -2e-05   Ek =1.9739e-05  Ep =-3.9478e-05
VERLET
Angular momentum: 39.4784*m
Step: 1000000  E = -2e-05   Ek =1.9739e-05  Ep =-3.9478e-05

Energies for three-body system
Step: 1000000  E = -0.004   Ek = 0.0040197  Ep =-0.0077056
For 10m_j
Step: 1000000  E =  -0.04   Ek =  0.041006  Ep = -0.077361
For 1000m_j
Step: 100000   E =     -4   Ek =     3.314  Ep =    -6.944

Energies for multi-body system dt=1e-5 N=1e5
Step: 100000   E = -0.004   Ek = 0.0040429  Ep =-0.0084804
\end{lstlisting}
